Planck and the Extra-terrestrials\\ or what the new SI means for astronomy

### 2020-04-10 spg/main.tex
version 2

In the 120 years since Planck speculated about it, we have yet to make
contact with non-human or extra-terrestrial cultures, and ask them if
they indeed share Planck's views on units.  What {\em has} happened,
however, is that physical constants have turned out to have meanings
not even Planck in 1899 had imagined.  As we all know, the following
year Planck himself discovered that $h$ (formerly known as $b$) was
about quantization.  A few years later the formal relation between
mass and energy through $c^2$ turned out to be real.  Starting in the
1920s $(\hbox{Planck mass})^{-2}\times(\hbox{proton mass})^3$ was
discovered to be the mass scale of stars.  Later in the 20th century,
$2e/h$ and $h/e^2$ turned out to be macroscopically observable as
Josephson's constant and von~Klitzing's constant.  And just in the
last few years, $c^5/G$ has been observed to be the power-output scale
for merging black holes.  We can only wonder what more surprises await
us.

@@

\bibliographystyle{apsrev4-1}
\bibliography{refs.bib}

>>>>

To illustrate working with the new SI in astronomy, let us consider an
example that sounds like science fiction, but is in fact a serious
design study: the solar gravity lens
mission.\cite{2020arXiv200211871T} It envisions a telescope larger
than the solar system with the sun's gravitational field as its main
lens, for mapping the surfaces of extra-solar planets.  The essential
physical effect is that light rays going past the sun with impact
parameter $R$ are gravitationally deflected by an angle
$4GM_\odot/(c^2 R)$.  In particular, the deflection of light from a
star that is barely covered by the sun will make that star appear to
be just outside the rim of the sun.  The deflection at the rim of the
sun is well known to be $1.75\unit{arcsec}$ and its first measurement
a century ago is the stuff of legend.\cite{Kennefick2012} Knowing the
solar mass in gravity-seconds and approximating the solar radius as
$R_\odot\approx c\times2\unit{s}$, we can trivially work out the
gravitational deflection angle $\alpha=4GM_\odot/(c^2 R_\odot)
\approx10^{-5}\unit{radians}$, which is indeed a little under
$2\unit{arcsec}$.  If we go far enough from the sun that its apparent
radius on the sky becomes smaller than the deflection angle, a
remarkable phenomenon will present itself: light sources precisely
behind the sun will appear as luminous rings (known as Einstein rings)
around the sun.  The threshold distance to see an Einstein ring is
$R_\odot/\alpha\approx c\times2\times10^5\unit{s}$ or $\approx
400\unit{au}$.  (Actually about $550\unit{au}$ with less drastic
approximations.)  The advantage of an Einstein ring is that it
enormously magnifies the source, so much so that a $1\unit{m}$
telescope located at the right place in deep space could resolve
features down to $10\unit{km}$ on an exoplanet at 30~parsecs.  That
level of detail is comparable to what the Hubble Space Telescope can
see on Mars.  The image-reconstruction problem involved would be a
major challenge in its own right, because the magnification produced
by Einstein rings is highly anisotropic, but simulations indicate that
it is solvable and that images could indeed be obtained.  The task
that remains, then, is to fly a fleet of space telescopes to
appropriate locations beyond $500\unit{au}$.  And within one lifetime,
if possible, please.  The currently most promising strategy for doing
this turns out to be (who would have guessed?) solar sails.  Solar
sails also provide a nice example of how kilograms and gravity-seconds
can co-exist.  Radiation pressure from the sun is, like gravity,
inverse squared in the distance.  Hence the acceleration on a
spacecraft with a solar sail will take the form
$$ \frac{S\,F_\odot}{r^2} - \frac{GM_\odot}{r^2} $$ where $r$ is the
distance from the sun, $F_\odot/r^2$ is the radiation pressure from
the sun, and $S$ is the effective sail area per unit mass.  For
simplicity we are neglecting the gravity of the planets and further
non-gravitational perturbations.  The solar-sail term in effect
reduces the gravitating mass of the sun, and could even reverse its
sign if $S$ is large enough.  Let us consider both terms at
$r=r_\oplus=1\unit{au}$.  Knowing the solar mass in gravity-seconds we
easily get $GM_\odot/r_\oplus^2\approx6\unit{mm}\unit{s}^{-2}$.  To
estimate $F_\odot/r_\oplus^2$ a convenient way is to recall the solar
constant, which is the energy flux from the Sun at the top of the
Earth's atmosphere.  This energy flux is, of course, in photons, and
the energy of a photon is $c$ times its momentum.  Holding up a mirror
normal to sunlight reflects the photons and imparts twice their
momentum to the mirror.  Hence, multiplying the solar constant by
$2/c$ gives the pressure on an optimally-aligned solar sail
$1\unit{au}$ from the sun.  The value of the solar constant
$1361\unit{W}\unit{m}^{-2}$ with small variations which we will
disregard here, and the corresponding pressure will be a little under
$10^{-5}\unit{Pa}$.  Comparing the terms, we see that we would need
$S\approx 600\unit{m}^2\unit{kg}^{-1}$ to make the sun a net repeller.
This may or may not be technologically feasible, but it is actually
not necessary, thanks to a further cunning plan.  This is to first
plunge the spacecraft in a very eccentric orbit close to the sun, and
then at perihelion deploy the solar sail.  Without the solar sail, the
spacecraft would simply have returned to repeat the orbit, but now it
can sail out of the solar system on its mission to the focal region of
the solar gravity lens.

Even with such a daring space mission, it will probably be some time
before we make contact with non-human or extra-terrestrial cultures,
so that we can ask them if they indeed share Planck's views on units.
Yet in the 120 years since Planck drew special attention to physical
constants, they have turned out to have meanings not even Planck in
1899 had imagined.  As we all know, the following year Planck himself
discovered that $h$ (formerly known as $b$) was about quantization.  A
few years later the formal relation between mass and energy through
$c^2$ turned out to be real.  Starting in the 1920s
$(hc/G)^{3/2}\times(\hbox{proton mass})^{-2}$ was discovered to be the
mass scale of stars.  Later in the 20th century, $2e/h$ and $h/e^2$
turned out to be macroscopically observable as Josephson's constant
and von~Klitzing's constant.  And just in the last few years, $c^5/G$
has been observed to be the power-output scale for merging black
holes.  We can only wonder what more surprises await us.

@@

\bibliographystyle{apsrev4-1}
\bibliography{main.bbl}
%\bibliography{refs.bib}
